A forest is more than its trees: haplotypes and inferred ARGs

### 2024-10-10 paper/writeup.tex
plot fixes
writing!

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}}
\newcommand{\tn}{\textnormal}
\newcommand{\ov}{\overline}
\newcommand{\tskit}{\texttt{tskit}}
\newcommand{\comment}[1]{{\color{violet} \it #1}}
\newcommand{\argmax}{\operatorname{argmax}}

@@

% NOTE: these are in alphabetical order for now; we can revisit that ordering
\author[$\dagger$]{Ava Bamforth}
\author[$\dagger$]{Halley Fritze}
\author[$\ddagger$]{Jerome Kelleher}
\author[$\dagger$]{Nathaniel Pope}
\author[$\ast$,$\dagger$,1]{Peter Ralph}

\affil[$\ast$]{Institute of Evolution and Ecology and Department of Biology, University of Oregon, Eugene, Oregon}
\affil[$\dagger$]{Department of Mathematics, University of Oregon, Eugene, Oregon}
\affil[$\ddagger$]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford}

@@

There has been substantial recent progress
in the problem of ``ARG inference'',
which seeks to infer (portions of) the ``ancestral recombination graph'' (or, ARG)
that describes how a set of genotypes samples are related to each other
at each position of the genome.
For reviews of this progress, see CITE CITE CITE.
Often such methods focus on the sequence of ``marginal trees'',
i.e., the genealogical trees that describe how each portion of the genome
was inherited by the focal genomes.
For instance, the ``succinct tree sequence'',
introduced by \citet{kelleher2016efficient},
is a common format for describing these inferred ARGs,
and is seeing wide use thanks in part to its efficiency and accompanying reliable toolkit,
\tskit (CITE).

However, the ARG is emphatically not merely a sequence of trees:
since each segment of an ancestor's genome that has been inherited by the focal genomes
may extend over a portion of the genome over which there are many distinct trees,
a single ancestor should usually be represented by nodes in many distinct trees.
In other words, one way to view an ARG is as a sequence of trees in which nodes may be shared across many trees,
but another way to view the ARG is as a set of relationships between segments of ancestral haplotypes.

Another reason we tend to focus on the trees is that
much of our intuition about inference of relationships from genomic data
comes from phylogenetics.
Indeed, all methods might very roughly be summarized as
``more similar sequences are more closely related''.
For instance, two sequences that share a drerived mutation
are (probably) more closely related over some span of genome surrounding the location where the mutation occurs.
\citet{fisher} pointed out that not only mutations
but also the ``junctions'' between distinct haplotypes,
if they could be somehow identified,
would be inherited as Mendelian markers.

@@

Consider the (small portion) of a hypothetical tree sequence in Figure~\ref{fig:conceptual}A.
On the first portion of the genome (left-hand tree), the sample nodes (labeled 0, 1, and 2)
coalesce into a small subtree: 1 and 2 find a common ancestor in ancestral node 3,
which finds a common ancestor with node 0 in ancestral node 4.
On the next portion of the genome (right-hand tree), sample node 2 has a different ancestor.
This seems reasonable, and a method that infers trees separately on each portion of the genome
could not be expected to produce anything different.
However, things change once we think about what this implies about haplotype inheritance.
Figure~\ref{fig:conceptual}B shows the implied inheritance of haplotypes,
with the haplotypes carried by 4 to the left and right of the recombination breakpoint labeled $L$ and $R$.
Here, sample node 2 has inherited the chunk of haplotype labeled $L$ from ancestral node 4 via 3,
and the haplotype to the right of this from some other node (and so doesn't carry haplotype $R$).
On the other hand, sample node 1 has inherited \emph{both} haplotypes $L$ and $R$
from ancestral node 4, but the trees imply that only haplotype $L$ is inherited via ancestral node 3.
This implies -- if taken literally -- that there must have been a recombination event
at some point between node 1 and node 4 that separated the $L$ and $R$ haplotypes,
and then these two ancestral (and nonoverlapping) haplotypes coalesced together in ancestral node 4.
Although this is possible, it seems unlikely --
a more parsimonious explanation is depicted in Figure~\ref{fig:conceptual}C,
in which sample node 1 inherits the entire $LR$ haplotype from ancestral node 4 through node 3
(and there is a recombination somewhere between node 3 and node 2).
This implies that ancestral node 3 inherits from node 4 on the right-hand tree as well,
which is depicted in Figure~\ref{fig:conceptual}D --
and so node 3 has become unary in this tree.
Note that the more parsimonious tree sequence also includes fewer edges:
the three distinct edges $4 \to 3$, $3 \to 1$, and $4 \to 1$ in Figure~\ref{fig:conceptual}B
have been reduced to the two edges
$4 \to 3$ and $3 \to 1$ in Figure~\ref{fig:conceptual}D.

@@

\subsection{Extend Haplotypes}

% * description (HALLEY)
% * proof of something:
%     - arrives at a local minimum?
%     - guess at typical reduction? (reduces edges by 1/3?)
    We define a \textit{tree sequence}, denoted $\T$, 
    as a tuple $\left(T_1,...,T_{|\T|}\right)$ of $|\T|$ trees with a sequence of points
	$0 = a_0 \leq a_1 \leq \cdots < a_{|\T|} = L$.
     Each $T_k$ is a tree describing genealogical relationships between samples
     over a distinct portion of the genome $[a_{k-1}, a_k)$.
    Given any tree sequence, our goal is to
    identify areas of implied inheritance of haplotypes.
    We do so by identifying paths in pairs of consecutive tree sequences
    and merging them, given the correct condition.
    For example, \figref{fig:extending_diagram}, if there exists paths neighboring trees which contain 
    identical ancestor (parent) and descendant (child) nodes, 
    we assume that there is implied coalescence of haplotypes between nodes, 
    and we will extend the path's edges to the neighboring tree.
     More generally, see \algorithmref{alg:hap}, for a pair of trees $T_k, T_{k+1}\in\mathbb{T}$,
     to transition from tree $T_k$ to tree $T_{k+1}$ we identify the 
     subforest of edges in $T_k$ that are removed to construct $T_{k+1}$
     and the subforest of edges in $T_{k+1}$ which were added to construct 
     $T_{k+1}$ from $T_k$. We call these subforests the out-forest and in-forest, respectively,
     and denote them as $F_O$ and $F_O$.
     For each edge $e \in F_O$ with child $c$,
     we construct a path $p_c^{F_O}$ comprised of edges in the $F_O$ subforest.
     If there exists an edge $e'\in F_I$ also with child $c$,
     we construct path $p_c^{F_I}$ from edges in the $F_I$ subforest.
     If there is a common node in both paths excluding $c$, ie. $(p_c^{F_O}\cap p_c^{F_I})\backslash \{c\}\neq\emptyset$, then there is a mergable path between both subforests. The merged path $\bar{p}_c = \hat{p}_c^{F_I} \cup \hat{p}^{F_O}$ is the union of both paths truncated ($\hat{p}$) at the common intersection node. We then insert $\bar{p}$ into $T_2$, postponing or deleting edges in its original in-forest $F_I$, see \figref{complicated_extending_diagram}. We perform this action across all consecutive pairs of tree sequences, going both forward and backward along the genome, until there are no more edges extended in this way.

\begin{figure}[!ht]
\begin{center}
	\includegraphics[width=5in]{edge_extend_method.pdf}
\end{center}
\caption{\comment{Change figure to say Extend halplotypes}A visualization of the \textit{extend haplotypes} method. Given two neighboring trees $T_1$ and $T_2$, $F_O$ in the tree $T_1$ contains the edge $3\to 0$. Starting at $0$, there is a path of edges $0 \to 3 \to 4$. The tree $T_2$ has edge $4 \to 0$ in $F_I$, and so there exists path $0 to 4$. These two paths share common end points, $0$ and $4$, and so we will extend path $0 \to 3 \to 4$ onto tree $T_2$, hence removing the edge $4\to 0$.
    \label{fig:extending_diagram}
}
\end{figure}

@@

	\caption{Extends haplotypes across a tree sequence $TS$.}
	\label{alg:hap}

%% Can most likely delete the extend-edges algorithm %%
%\begin{algorithm2e}[!ht] 
%	
%\SetStartEndCondition{ }{}{}%
%\SetKwProg{Fn}{def}{\string:}{}
%\SetKwFunction{Range}{range}%%
%\SetKw{KwTo}{in}
%\SetKwFor{For}{for}{\string:}{}%
%\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
%\SetKwFor{While}{while}{:}{fintq}%
%\SetKw{Break}{break}
%\newcommand\forcond{$i$ \KwTo\Range{$n$}}
%\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
%
%\caption{Extends edges across a tree sequence $TS$.}
%\label{alg:edge}
%
%\SetKwFunction{ex}{Extend Edges}
%\SetKwFunction{fwd}{Extend}
%\DontPrintSemicolon
%\LinesNotNumbered
%%\SetAlgoNoEnd
%\input{A tree sequence $\T$}
%\output{New tree sequence,$\T'$}
%
%\BlankLine 
%
%\Fn{\fwd (tree sequence $\T$, forwards = \tn{True})}{
%{	
%\eIf{forwards=\tn{True}}{$\T = \left(T_1,...,T_{|\T|}\right)$\BlankLine}
%{\BlankLine
%	$\T = \left(T_{|\T|},..., T_1\right)$\;}
%}
%	\For{$T_k\in \T$}{
%	Find all edge pairs $(e_1,e_2)$ such that $e_1 = a\to b$ and $e_2 = b\to c$ for some nodes $a,b,c$.\;
%	\For{edges $f$ in $T_{k+1}$}{
%		\If{$f = a \to c$ and $b$ is not in $T_{k+1}$}{
%		Remove $f$ from $T_{k+1}$.\;
%		Add $e_1$ and $e_2$ to $T_{k+1}$.\;
%		Update $\T$.\;
%	}
%	}
%}\;						
%\KwRet{$\T$}\;
%}
%\BlankLine						
%\Fn{\ex ($\T$, iter=100)}{
%	\For{$*$ \KwTo \Range(iter)}{
%		$\T' \gets \T$.\fwd(fowards=\tn{True})\;
%		$\T' \gets \T$.\fwd(fowards=\tn{False})\;
%		\If{Number of edges in $\T$ = Number of edges in $\T'$}{
%			\Break\BlankLine}
%		}
%}
%\KwRet{$\T'$}\;
%\end{algorithm2e}

@@

\subsection{Discrepancy Function}
% * How to measure agreement that includes haplotypes
% * Definition and Algorithm
% * Supp fig: runtime ~ # trees, samples
% * fig: how add edges reduces discrepancy (RESULTS)
%       - compare to tsinfer
% * Show: how add edges reduces discrepancy (RESULTS)
% * THINK of a way to asynchronously computing discrepancy (???)

\par \comment{This needs work. Choose $n_1$ or $i$ for node not both. Do you want it through the perspective of TS or through nodes? You wording is mixed on this}
 Now with an algorithm in place
 we need a metric to discern
 if trees modified with extend edges
 in fact correctly infer haplotypes in our original tree sequence.
  % Robinson-Foulds Paper: https://www.sciencedirect.com/science/article/abs/pii/0025556481900432?via%3Dihub 
 An immediate choice of metric would be the Robinson-Foulds metric \comment{CITE} which compares nodes between individual pairs of trees.
 However, in an effort to infer coalescent haplotypes, our function chooses to compare nodes
 across the entire genome. 
 \comment{something something something...}

>>>>

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}}
\newcommand{\ind}{\mathbf{1}}
\newcommand{\tn}{\textnormal}
\newcommand{\ov}{\overline}
\newcommand{\tskit}{\texttt{tskit}}
\newcommand{\argmax}{\operatorname{argmax}}
\newcommand{\dis}{\operatorname{dis}}
\newcommand{\agree}{\operatorname{agree}}
\newcommand{\comment}[1]{{\color{violet} \it #1}}

@@

% NOTE: these are in alphabetical order for now; we can revisit that ordering
\author[$\dagger$]{Ava Bamforth}
\author[$\dagger$]{Halley Fritze}
\author[$\ddagger$]{Jerome Kelleher}
\author[$\dagger$]{Nathaniel Pope}
\author[$\ast$,$\dagger$,S]{Peter Ralph}

\affil[$\ast$]{Institute of Evolution and Ecology and Department of Biology, University of Oregon, Eugene, Oregon}
\affil[$\dagger$]{Department of Mathematics, University of Oregon, Eugene, Oregon}
\affil[$\ddagger$]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford}
\affil[$\S$]{Department of Data Science, University of Oregon, Eugene, Oregon}

@@

There has been substantial recent progress
in the problem of ``ARG inference'',
which seeks to infer (portions of) the ``ancestral recombination graph'' (or, ARG)
that describes how a set of genotypes samples are related to each other
at each position of the genome \citep{XXX}.
One way of viewing the ARG is as a sequence of ``marginal trees'',
i.e., the genealogical trees that describe how each portion of the genome
was inherited by the focal genomes.
This is reflected in methodology of some ARG inference methods \citep[e.g.,][]{Relate},
in metrics used to assess inference accuracy \citep{XXX},
as well as in basic terminology.
For instance, 
the ``succinct tree sequence'',
introduced by \citet{kelleher2016efficient},
is a common format for describing these inferred ARGs,
and is seeing wide use thanks in part to its efficiency and accompanying reliable toolkit,
\tskit \citep{XXX}.

However, the ARG is emphatically not merely a sequence of trees:
viewed another way, it describes inheritance relationships between ancestral haplotypes.
These two points of view are related because a single haplotype
may extend over many marginal trees;
in other words, the internal nodes in the trees are labeled, and many of these labels
are shared between adjacent trees.

Another reason we tend to focus on the trees is that
much of our intuition about inference of relationships from genomic data
comes from phylogenetics.
Indeed, all methods might very roughly be summarized as
``more similar sequences are more closely related''.
For instance, two sequences that share a drerived mutation
are (probably) more closely related over some span of genome surrounding the location where the mutation occurs.
It has long been observed 
that not only mutations
but also the ``junctions'' between distinct haplotypes,
if they could be somehow identified,
would be inherited as Mendelian markers
\citet{fisher1954fuller,chapman2003model}.
In more modern terminology, 
even in the absence of new mutations,
recombination between distinct haplotypes can create a novel haplotype
whose relationships and origination time could be inferred.

Haplotype identity has been relatively overlooked in the literature on ARG inference so far --
all metrics that have been used so far to measure accuracy of inferred ARGs
depend only on the sequence of marginal trees,
not on sharing of ancestral haplotype identities between them.
Suppose two methods are used to infer ARGs from the same set of genotyped samples.
Although both ARGs describes relationships between many haplotypes,
only those haplotypes that represent the focal set of genotyped samples
can be unambiguously said to represent the same thing in both.
So, most metrics only describe these relationships,
which ignores the additional information carried by the structure of other haplotypes.
For instance, \citet{XXX} compared true and inferred ARGs
using average XXX distance between trees across a regular sequence of genomic positions,
using sampled genotypes as labels,
while \citet{brandt2022evaluation} compared times to most recent common ancestor
between pairs of sampled genomes.
Neither is affected by shared haplotype structure -
two ARGs could be identiical by either measure
but have completely different identity by descent (IBD) segment structure
(as defined in the context of the ARG, \citep{XXX}).

In this paper, we study various aspects of haplotype identity in ARGs.
First, we describe a deterministic algorithm that
extends the span over which ancestral haplotypes extend
using the principle that intermediate nodes in inheritance paths
should remain unchanged when possible.
These extended ancestral haplotypes manifest as unary nodes in the marginal trees.
To quantify how accurate the new information is,
we define and describe how to compute new measures of (dis-)agreement between ARGs
that are motivated by the XXX distance between trees
but account for haplotype identity.
These measures show that the vast majority of these extended haplotypes are correct,
and that substantial information about haplotypes is contained in these nodes
in inferred trees as well.

@@

Consider the (small portion) of a hypothetical tree sequence in \figref{fig:conceptual}A.
On the first portion of the genome (left-hand tree), the sample nodes (labeled 0, 1, and 2)
coalesce into a small subtree: 1 and 2 find a common ancestor in ancestral node 3,
which finds a common ancestor with node 0 in ancestral node 4.
On the next portion of the genome (right-hand tree), sample node 2 has a different ancestor.
This seems reasonable, and a method that infers trees separately on each portion of the genome
could not be expected to produce anything different.
However, things change once we think about what this implies about haplotype inheritance.
\figref{fig:conceptual}B shows the implied inheritance of haplotypes,
with the haplotypes carried by 4 to the left and right of the recombination breakpoint labeled $L$ and $R$.
Here, sample node 2 has inherited the chunk of haplotype labeled $L$ from ancestral node 4 via 3,
and the haplotype to the right of this from some other node (and so doesn't carry haplotype $R$).
On the other hand, sample node 1 has inherited \emph{both} haplotypes $L$ and $R$
from ancestral node 4, but the trees imply that only haplotype $L$ is inherited via ancestral node 3.
This implies -- if taken literally -- that there must have been a recombination event
at some point between node 1 and node 4 that separated the $L$ and $R$ haplotypes,
and then these two ancestral (and nonoverlapping) haplotypes coalesced together in ancestral node 4.
Although this is possible, it seems unlikely --
a more parsimonious explanation is depicted in \figref{fig:conceptual}C,
in which sample node 1 inherits the entire $LR$ haplotype from ancestral node 4 through node 3
(and there is a recombination somewhere between node 3 and node 2).
This implies that ancestral node 3 inherits from node 4 on the right-hand tree as well,
which is depicted in \figref{fig:conceptual}D --
and so node 3 has become unary in this tree.
Note that the more parsimonious tree sequence also includes fewer edges:
the three distinct edges $4 \to 3$, $3 \to 1$, and $4 \to 1$ in \figref{fig:conceptual}B
have been reduced to the two edges
$4 \to 3$ and $3 \to 1$ in \figref{fig:conceptual}D.

So, given the ARG shown in \figref{fig:conceptual}A\&B,
it should be possible to extend the ancestral haplotype represented by node 3
to obtain the ARG shown in \figref{fig:conceptual}C\&D,
thus adding additional information to the ARG.
This might be surprising,
as intuition from phylogenetics suggests we can only infer
information about the branching points in the tree, not intermediate (unary) nodes.
The goal of this paper is to answer:
How can we do this, and how accurate is the resulting inference?

@@

\subsection{Notation and terminology}

We use work with the \emph{succinct tree sequence} representation of the ARG (henceforth, ``tree sequence''),
to take advantage of the extensive, stable, and well-documented set of open-source tools
available in \tskit{} \citep{tskit}.
So, our terminology and notation follows \citet{ralph2020efficiently}.
For our purposes here,
a tree sequence $\T = (N, E)$ contains a set of \emph{nodes} $N$ 
which represent ancestral segments of genome,
and \emph{edges} $E$ which represent relationships between nodes over different regions of the genome.
Each node $n \in N$ has a \emph{time} $t_n$,
which is the amount of time in the past that the individual who carried that segment of genome lived.
Each edge $e \in E$ describes inheritance between a parent node $p_e$ and child node $c_e$,
over a segment of genome $[\ell_e, r_e)$.
(The tree sequence includes substantially more information, including genotypes, but we will not use that in this paper.)
Suppose that the unique elements of the set of left and right edge endpoints
are $0 = a_0 < a_1 < \cdots < a_{n} = L$, where $L$ is the length of the genome.
Using this information, one can construct the sequence of 
trees $\left(T_1,...,T_{n}\right)$ that describe how the nodes are related to each other along the genome:
each $T_k$ is a tree whose nodes are in $N$
and that describes relationships on the half-open interval $[a_{k-1}, a_k)$.
Not all nodes appear in each tree,
and we say $n \in T_k$ for a node $n$ if the tree $T_k$ describes at least one parent-child relationship
for node $n$.

\subsection{An algorithm to extend haplotypes}

% Description of algorithm
Given a tree sequence, our goal is to
identify areas of implied inheritance of haplotypes.
Generalizing from \figref{fig:conceptual} (or \figref{fig:extending_diagram}?),
we do this by identifying paths of inheritance that are shared across a sequence of marginal trees
but for which some of the intermediate nodes are missing.
Concretely, suppose that 
if in tree $T_k$ there is a chain of inheritance
$p \to u_1 \to \cdots \to u_m \to c$
(where $a \to b$ denotes a parent-child relationship)
and in tree $T_{k+1}$ there is a chain of inheritance
$p \to v_1 \to \cdots \to v_n \to c$,
where $\{u_i\}_{i=1}^m$ and $\{v_j\}_{j=1}^n$ are disjoint.
This situation implies that $c$ inherited from $p$ over the entire interval $[a_{k-1}, a_{k+1})$,
so it seems reasonable to assume that $c$ has inherited from $p$ \emph{along the same path} for that entire interval.
In other words, the intermediate nodes $\{u_i\}$ should also lie on the path from $c$ to $p$ in tree $T_{k+1}$,
and conversely the nodes $\{v_j\}$ should lie on that path in tree $T_k$.
Of course, this does not always make sense --
for instance, if $u_i$ is already represented somewhere else in $T_{k+1}$,
or if $t_{u_i} = t_{v_j}$ for some $i$ and $j$.
for which the intermediate nodes are not in the other tree
So, we restrict our attention to pairs of such paths in adjacent trees
for which
$u_i \notin T_{k+1}$ for all $1 \le i \le m$,
$v_j \notin T_k$ for all $1 \le j \le n$,
and the times of the nodes $\{u_i\}$ and $\{v_j\}$ are unique.
Call a pair of such paths \emph{mergeable}.
So, the goal of our algorithm is to iterate over trees,
identify mergeable pairs of paths,
and then extend the nodes $\{u_i\}$ to $T_{k+1}$.
(We also extend $\{v_j\}$ to $T_k$, but on a backwards pass.)

\begin{figure}[!ht]
\begin{center}
	\includegraphics[width=5in]{edge_extend_method.pdf}
\end{center}
\caption{\comment{Change figure to say Extend halplotypes} A visualization of the \textit{extend haplotypes} method. Given two neighboring trees $T_1$ and $T_2$, $F_O$ in the tree $T_1$ contains the edge $3\to 0$. Starting at $0$, there is a path of edges $0 \to 3 \to 4$. The tree $T_2$ has edge $4 \to 0$ in $F_I$, and so there exists path $0 to 4$. These two paths share common end points, $0$ and $4$, and so we will extend path $0 \to 3 \to 4$ onto tree $T_2$, hence removing the edge $4\to 0$.
    \label{fig:extending_diagram}
}
\end{figure}

@@

A relatively efficient algorithm to do this is described in \algorithmref{alg:extend}, 
and implemented in \tskit{} as \texttt{extend\_haplotypes}.
The algorithm considers each tree transition from $T_k$ to $T_{k+1}$ in turn, updating its internal state
(which includes possibly modifying $T_k$) as it goes.
Suppose we are at the transition from tree $T_k$ to tree $T_{k+1}$,
which is done by first removing a set of edges $O$
and then adding another set of edges $I$.
$O$ defines a sub-forest $F_O$ of $T_k$,
and $I$ defines a sub-forest $F_I$ of $T_{k+1}$.
The key step in the algorithm is to determine whether the pair of paths
that terminate in a given node in two adjacent trees are mergeable,
\algorithmref{alg:mergeable}, which works as follows.
If a pair of paths is mergeable,
then the edges of the two paths must lie in $O$ and $I$, respectively.
Suppose an edge in $O$ has child $c$.
To see if $c$ is the base of a pair of mergeable paths,
the algorithm traverses up from $c$ in both $F_O$ and $F_I$;
terminating if a node in the other tree is found
(i.e., if the node traversed in $F_O$ is in $T_{k+1}$ or if the node traversed in $F_I$ is in $T_k$)
or if a pair of traversed nodes have the same time.
If these two traversals end in the same node $p$, the paths are mergeable.
Iterating over all edges in $O$ will thus find all mergeable pairs of paths.
There is often more than one pair of mergeable paths in a tree transition;
so, the algorithm merges pairs of mergeable paths,
starting with pairs that add the smallest number of new edges,
until no more are found.
\comment{Maybe we need to highlight in a figure how new edges can be added.}

\algorithmref{alg:extend} simplifies the algorithm in several ways --
for instance, the bookeeping required to keep track of $T_k$ and $T_{k+1}$ is omitted.
Furthermore, as described the algorithm does one left-to-right pass over the tree sequence;
in practice we do repeated passes in both directions until no changes can be made.
The main step that is omitted is a description of the \texttt{merge} operation,
which performs the actual extending of haplotypes.
This algorithm is essentially the same as \texttt{mergeable} in \algorithmref{alg:mergeable},
except with additional bookkeeping.
Roughly speaking, the algorithm
traverses up from the shared base node $c$,
doing the appropriate operations to insert the nodes along the path found in $T_k$
into the path in $T_{k+1}$.
To do this, some edges that end at $a_k$ will be extended to end at $a_{k+1}$;
some edges that begin at $a_k$ will be postponed to begin at $a_{k+1}$,
and some entirely new edges may be added.
Furthermore, the trees $T_k$ and $T_{k+1}$ (and corresponding forests $F_O$ and $F_I$)
need to be updated.

\begin{algorithm2e}[!ht]
	\SetStartEndCondition{ }{}{}%
	\SetKwProg{Fn}{def}{\string:}{}
	\SetKwFunction{Range}{range}%%
	\SetKw{KwTo}{in}
	\SetKwFor{For}{for}{\string:}{}%
	\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
	\SetKwFor{While}{while}{:}{fintq}%
	\SetKw{Break}{break}
	\newcommand\forcond{$i$ \KwTo\Range{$n$}}
	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
    \caption{
        Given a node $c$, trees $T_O$ and $T_I$,
        and sub-forests $F_O$ and $F_I$
        such that removing $F_O$ and adding $F_I$ turns $T_O$ into $T_I$,
        checks to see if the paths upwards from $c$ in $T_O$ and $T_I$ are mergeable.
        If the paths are mergeable then this returns the number of new edges
        that would be added by extending the path from $T_O$ to $T_I$;
        otherwise, returns $\infty$.
        Let $P_O[n]$ and $P_I[n]$ be the parents of node $n$
        in the set of edges to be removed and added, respectively
        (i.e., in $F_O$ and $F_I$).
    }
	\label{alg:mergeable}
	\SetKwFunction{mergeable}{Mergeable}
	\DontPrintSemicolon
    \Fn{\mergeable($c$, $T_O$, $T_I$)}{
        $p_i = P_I[c]$\;
        $t_i = t[p_i]$\;
        $p_o = P_O[c]$\;
        $t_o = t[p_o]$\;
        $n_e = n = 0$\;
        \While{\tn{True}}{
            $y_i = (p_i \neq NULL)
                \text{ and } (p_i \notin T_O)
                \text{ and } (t_i < t_o)$ \;
            $y_o = (p_o \neq NULL)
                \text{ and } (p_o \notin T_I)
                \text{ and } (t_o < t_i)$ \;
            \If{not ($y_i$ or $y_o$)}{ \Break\; }
            \eIf{$y_i$}{
                \If{$P_I[c] \neq p_i$ and $P_O[c] \neq p_i$}{
                    $n_e += 1$\;
                }
                $c = p_i$\;
                $p_i = P_I[p_i]$\;
                $t_i = (p_i = NULL) ? \infty : t[p_i]$\;
            }{
                \If{$P_I[c] \neq p_o$ and $P_O[c] \neq p_o$}{
                    $n_e += 1$\;
                }
                $c = p_o$\;
                $p_o = P_O[p_o]$\;
                $t_o = (p_o = NULL) ? \infty : t[p_o]$\;
                $n += 1$\;
            }
        }
        \If{$n = 0$ or $p_i \neq p_o$ or $p_i = NULL$}{
            $n_e = \infty$\;
        }
        \KwRet{$n_e$}\;
    }
\end{algorithm2e}

\begin{algorithm2e}[!ht]
	\SetStartEndCondition{ }{}{}%
	\SetKwProg{Fn}{def}{\string:}{}
	\SetKwFunction{Range}{range}%%
	\SetKw{KwTo}{in}
	\SetKwFor{For}{for}{\string:}{}%
	\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
	\SetKwFor{While}{while}{:}{fintq}%
	\SetKw{Break}{break}
	\newcommand\forcond{$i$ \KwTo\Range{$n$}}
	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
    \caption{
        Extend haplotypes.
        \comment{TODO: include this in the previous algorithm?}
    }
	\label{alg:extend}
	\SetKwFunction{mergeable}{Mergeable}
	\SetKwFunction{merge}{Merge}
	\SetKwFunction{extend}{ExtendHaplotypes}
	\DontPrintSemicolon
    \Fn{\extend($\T$)}{
        \For{$k$ in $1 \ldots N$}{
            $M = 0$\;
            $M' = \infty$\;
            \While{$M < \infty$}{
                \For{$e \in O_k$}{
                    $m$ = \mergeable($c_e$, $T_k$, $T_{k+1}$)\;
                    \eIf{$m < M$}{
                        \merge($c_e$, $T_k$, $T_{k+1}$)\;
                    }{
                        $M' = \min(m, M)$\;
                    }
                }
                $M = M'$\;
                $M' = \infty$\;
            }
        }
    }
\end{algorithm2e}

Since the algorithm needs to take multiple passes over the tree sequence
in each direction,
an important practical quesiton for this algorithm is:
how passes do we need to do?
The algorithm is monotone (spans of ancestral nodes only increase),
so it is guaranteed to require only a finite number of passes,
but it is also not hard to construct pathalogical cases that require an arbitrary number of passes.
\figref{fig:num_passes} shows empirical results from simulated tree sequences,
which suggest that \comment{XXX TODO}.

\begin{figure}[!ht]
    \caption{
        Number of edges in the tree sequence (vertical axis)
        after each pass of the algorithm (number of passes on the horizontal axis),
        for tree sequences with different numbers of samples (different lines).
        \comment{TODO}
        \label{fig:num_passes}
    }
\end{figure}

@@

\subsection{Discrepancy Function}
% * How to measure agreement that includes haplotypes
% * Definition and Algorithm
% * Supp fig: runtime ~ # trees, samples
% * fig: how add edges reduces discrepancy (RESULTS)
%       - compare to tsinfer
% * Show: how add edges reduces discrepancy (RESULTS)

If we begin with a tree sequence containing unary nodes,
it is straightforward to remove the portions of each node's span on which it is unary,
apply \algorithmref{alg:extend},
and quantify how much node span was correctly or incorrectly added.
However, we are also interested in whether \algorithmref{alg:extend}
improves \emph{inferred} tree sequences.
Since we are not aware of any current methods for measuring (dis)agreement between tree sequences
that take into account haplotype identity,
we define a measure of \emph{discrepancy} to quantify this.

It is helpful to first describe what we compute in the simple case.
\figref{fig:node-spans}A shows the spans for each node in three tree sequences:
for each node, the top point is the total span in a ``true'' tree sequence
that was simulated with \texttt{msprime} \citep{msprime}
in a way that included spans over which nodes are unary in the output
\comment{(using which option?)}.
The bottom point is the span of the node after \emph{simplifying} the tree sequence
\citep{kelleher2018efficient},
which removes all unary nodes from all trees.
\comment{Usually (?)} between the two is the span of the node after applying \algorithmref{alg:extend}
to the simplified tree sequence.
So, the difference between the middle and bottom points is the amount by which
the span of that node has been extended.
\figref{fig:node-spans}B \comment{TODO} shows how much of this span is incorrect:
for each node, it gives the percent of the span after running \algorithmref{alg:extend}
that is incorrect (i.e., not present in the original tree sequence).
The values in these plots (perhaps summed over nodes)
therefore quantify how much the ancestral haplotypes represented by each node
were correctly, and incorrectly, extended.
They also tell us how much of the relationships in the original tree sequence
are represented in the result.

Now suppose that instead of comparing two tree sequences with the same set of nodes,
we wish to compare two tree sequence for which we know the sample nodes are the same
but are not given an identification between nodes otherwise
(as for instance with a true, simulated tree sequence and one inferred from its genotypes).
\comment{should have introduced ``samples'' before this}
Call the two tree sequence $T_1$ and $T_2$, which should have the same genome length
and the same set of sample nodes.
We would like to measure (a) how much of $T_1$ is found in $T_2$;
(b) how much of $T_1$ is \emph{not} found in $T_2$; and
(c) how much of $T_2$ is not found in $T_1$.
(Think of these three quantities as the size of the intersection
and two differences between the tree sequences,
thought of vaguely as sets.)
Roughly speaking, we first identify matching nodes
as those whose sets of descendant samples agree for the largest span along the genome,
and then compute for how much of their spans do their descendant samples agree (or not).

Concretely, suppose that the two tree sequences have the same set of breakpoints
between trees,
so that $T_1^{(1)}, \ldots, T_N^{(1)}$ are the trees in $\T_1$
and $T_1^{(2)}, \ldots, T_N^{(2)}$ are the trees in $\T_2$.
For a node $n$ and tree $T$
let $S(T, n)$ denote the set of samples that inherit from $n$ in $T$,
and for a pair of nodes $n_1$ and $n_2$ with $n_1$ in $\T_1$ and $n_2$ in $\T_2$,
define
\begin{align*}
    m(n_1, n_2)
    =
    \sum_{k=1}^N (a_k - a_{k-1}) \ind_{S(T^{(1)}_k(n_1) = S(T^{(2)}_k(n_2)} ,
\end{align*}
which is the total span over which the samples below $n_1$ in $\T_1$
matches the samples below $n_2$ in $\T_2$.
Also let
\begin{align*}
    s(\T, n) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{n \in T_k} ,
\end{align*}
the total span that node $n$ is present in the marginal trees.
The \emph{agreement} between $\T_1$ and $\T_2$ is then defined to be
\begin{align*}
    \agree(\T_1, \T_2)
    =
    \max_{\beta:N_1 \to N_2} \sum_{n \in N_1} m(n, \beta(n)) ,
\end{align*}
where the maximum is over all mappings nodes in $\T_1$ to nodes in $\T_2$.
(Note that multiple nodes in $\T_1$ may be mapped to the same node in $\T_2$,
and that there may be no nodes in $\T_1$ that are mapped to some nodes in $\T_2$.)
Since the maximum is independent over nodes, we may define for each node $n_1 \in \T_1$
it's \emph{best matching node} in $\T_2$ as
\begin{align*}
    \alpha(n_1) = \argmax_{n_2 \in N_2} m(n_1, n_2) ,
\end{align*}
so that
\begin{align*}
    \agree(\T_1, \T_2)
    =
    \sum_{n \in N_1} m(n, \alpha(n)) .
\end{align*}
If the best-matching node is unique, we define $\alpha(n_1)$ to be the node in $T_2$
out of those minimizing $m(n_1, n_2)$ that minimizes $|t^{(1)}_{n_1} - t^{(2)}_{n_2}|$
(and if \emph{this} is not unique, we pick an arbitrary one) --
however, this potential ambiguity does not affect the definition of $\agree(\T_1, \T_2)$.
We then define the \emph{discrepancy} of $\T_1$ from $\T_2$ by
\begin{align*}
    \dis(\T_1, \T_2)
    =
    \sum_{n \in N_1} (s(\T_1, n) - m(n, \alpha(n))) ,
\end{align*}
which is the total span of all nodes in $\T_1$
for which their descendant samples do \emph{not} match those of their best match in $\T_2$.
